Stability and accuracy of free surface time integration in viscous flows

### 2016-09-13 free-surface-paper.tex
Fix sign error.
Address some comments.

A large number of studies have approximated free surfaces in the interior of the domain by using the 
`sticky air' approximation (see \citet{crameri2012comparison} and references therein). In this approximation there is a low-viscosity, low-density layer in the fluid 
(termed `air', though its viscosity is much higher) above the free surface, effectively decoupling it from the boundary. 
Typically a free-slip boundary condition is used above the sticky air, though an open boundary may be preferable \citep{hillebrand2014using}.

Finally, one can use a true free surface, where a stress-free boundary condition is applied on the boundary 
of the domain. In this case, there can be flow in and out of the domain, so the boundary must move in time.
A true free surface has mathematical elegance in that the boundary condition of the domain more closely 
matches the boundary conditions which one is trying to model, but it typically requires a deformable 
domain with frequent remeshing to avoid ill-conditioned cells.

@@

Several studies have attempted to alleviate the timestepping requirements imposed by the sloshing 
instability. Since the most expensive part of geodynamic simulations is typically the Stokes solve,
most free surface calculations have preferred to use explicit timestepping methods for the free 
surface. \citet{kaus2010stabilization} proposed a quasi-implicit scheme which modifies the discretized 
Stokes matrix, giving it better stability properties.  \citet{popov2008slim3d}, \citet{kramer2012implicit}, and \citet{furuichi2015implicit}
explored methods for solving the transport of the free surface implicitly.

@@

We begin with the incompressible momentum conservation equations for creeping incompressible flow:
\begin{equation}
\begin{aligned}
\nabla \cdot \mathbf{T} &= \rho \mathbf{g} \\
\nabla \cdot \mathbf{u} &= 0,
\end{aligned}
\label{eq:momentum_mass_conservation}
\end{equation}
where $\mathbf{u}$ is the fluid velocity, $\rho$ is the fluid density, and $\mathbf{g}$ is the force due to gravity.
$\mathbf{T}$ is the stress tensor for a Newtonian fluid, defined by
\begin{equation}
\mathbf{T} = 2 \eta \varepsilon(\mathbf{u}) - p \mathbf{I},
\label{eq:stress_tensor}
\end{equation}
where $\eta$ is the viscosity, $\varepsilon(\mathbf{u}) = \frac{1}{2}(\nabla \mathbf{u} + (\nabla \mathbf{u} )^T )$
is the strain-rate tensor, and $\mathbf{I}$ is the identity tensor.
Substituting the stress tensor into Equation~\eqref{eq:momentum_mass_conservation} gets the familiar form of the Stokes equation:
\begin{equation}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}.
\label{eq:stokes}
\end{equation}

For the purposes of this analysis it is useful to define a hydrostatic reference state where the 
velocity $\mathbf{u}$ is zero:
\begin{equation}
- \nabla p_0 = \rho_0 \mathbf{g},
\label{eq:hydrostatic_stokes}
\end{equation}
where $p_0$ is the reference hydrostatic pressure and $\rho_0$ is a reference density profile (which may vary with depth).
The total pressure and density can then be decomposed into variations about their 
reference values: $\rho = \rho_0 + \rho^\prime$, $p = p_0 + p^\prime$.
Using this, the hydrostatic reference state (Equation~\eqref{eq:hydrostatic_stokes}) may be subtracted from Equation~\eqref{eq:stokes}.

This gives rise to the following time dependent, coupled system with unknowns $\mathbf{u}(t)$ and $\mathbf{p}(t)$:
\begin{equation}
\begin{aligned}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime &= \rho^\prime \mathbf{g} \qquad &\text{in } &\Omega(t)\\
\nabla \cdot \mathbf{u} &= 0  \qquad &\text{in } &\Omega(t)
\end{aligned}
\label{eq:final_system}
\end{equation}
defined on the bounded, moving domain $\Omega(t)\subset \mathbb{R}^d$ with boundary $\partial \Omega = \Gamma_0 \cup \Gamma_F$
split into fixed (Dirichlet) and free surface parts $\Gamma_0$ and $\Gamma_F$, respectively.
The boundary conditions are given by
\begin{equation}
\begin{aligned}
\mathbf{u} &= 0 &\qquad \text{on } &\Gamma_0 \\
\mathbf{T}\cdot \mathbf{n} &= 0 &\qquad \text{on } &\Gamma_F.
\end{aligned}
\end{equation}
For the sake of economy, we neglect inhomogeneous stress boundary conditions, though it would be straightforward to include them.
We also defer discussion of free slip boundary conditions.
The domain at time $t$ is defined by advecting a reference configuration $\Omega_0$ by the domain velocity $\mathbf{u}(t)$
\begin{equation}
 \Omega(t) = \Omega_0 + \int_{t_0}^t \mathbf{u}(t) \text{ d}t.
 \label{eq:domain_evolution}
\end{equation}
We defer the discussion of the velocity of the discretized domain to Section~\ref{sec:remeshing}.
Finally, we denote the displacement at time $t$ by $\zeta = \int_{t_0}^t \mathbf{u}_\mathrm{surface}(t)\cdot \mathbf{n} \text{ d}t$, leading to the evolution
equation
\begin{equation}
\frac{\text{d} \zeta}{\text{d}t} = \mathbf{u \cdot \mathbf{n}} \quad \textrm{on  }  \Gamma_F.
\label{eq:surface_evolution}
\end{equation}

@@

\section{Conclusion}
We have analyzed stability of free surface boundary conditions in geodynamic simulations and 
demonstrated the cause of sloshing instabilities using a normal mode analysis.
This perspective on the problem allowed us to construct an explicit finite difference 
scheme which is first order accurate in time and is unconditionally stable.
The nonstandard finite difference scheme is simple to implement, and 
requires no modifications to the system matrix.

>>>>

A large number of studies have approximated free surfaces in the interior of the domain by using the 
`sticky air' approximation (\citet{matsumoto1983numerical}; for a comprehensive review see \citet{crameri2012comparison} and references therein). In this approximation there is a low-viscosity, low-density layer in the fluid 
(termed `air', though its viscosity is much higher) above the free surface, effectively decoupling it from the boundary. 
Typically a free-slip boundary condition is used above the sticky air, though an open boundary may be preferable \citep{hillebrand2014using}.

Finally, one can use a true free surface, where a stress-free boundary condition is applied on the surface of the simulation. 
With this setup there can be flow in and out of the free surface, so it must move in time.
In many cases this is accomplished through a Lagrangian or arbitrary Lagrangain-Eulerian (ALE) scheme (see Section~\ref{sec:implementation}).
\citet{duretz2016free} proposed an alternative scheme based on using an Eulerian staggered finite difference mesh
with a modified stencil at the surface to capture the free surface boundary condition.

A true free surface has mathematical elegance in that the boundary condition of the domain more closely 
matches the boundary conditions which one is trying to model, but it typically requires a deformable 
domain with frequent remeshing to avoid ill-conditioned cells. 
In more complex geodynamic simulations there may be portions of the surface which
correspond to regions below air (i.e. continental crust) and other portions that lie below water (i.e. oceanic crust).
In these cases the density contrast at the surface will vary, requiring variations
in the pressure boundary conditions (though viscous stresses should still be zero).

@@

Several studies have attempted to alleviate the timestepping requirements imposed by the drunken sailor 
instability. Since the most expensive part of geodynamic simulations is typically the Stokes solve,
most free surface calculations have preferred to use explicit timestepping methods for the free 
surface. \citet{kaus2010stabilization} proposed a quasi-implicit scheme which modifies the discretized 
Stokes matrix, giving it better stability properties.  \citet{popov2008slim3d}, \citet{kramer2012implicit}, and \citet{furuichi2015implicit}
explored methods for solving the transport of the free surface implicitly.

@@

We begin with the incompressible momentum conservation equations for creeping incompressible flow:
\begin{equation}
\begin{aligned}
\nabla \cdot \mathbf{T} &= - \rho \mathbf{g} \\
\nabla \cdot \mathbf{u} &= 0,
\end{aligned}
\label{eq:momentum_mass_conservation}
\end{equation}
where $\mathbf{u}$ is the fluid velocity, $\rho$ is the fluid density, and $\mathbf{g}$ is the force due to gravity.
$\mathbf{T}$ is the stress tensor for a Newtonian fluid, defined by
\begin{equation}
\mathbf{T} = 2 \eta \varepsilon(\mathbf{u}) - p \mathbf{I},
\label{eq:stress_tensor}
\end{equation}
where $\eta$ is the viscosity, $\varepsilon(\mathbf{u}) = \frac{1}{2}(\nabla \mathbf{u} + (\nabla \mathbf{u} )^T )$
is the strain-rate tensor, and $\mathbf{I}$ is the identity tensor.
Substituting the stress tensor into Equation~\eqref{eq:momentum_mass_conservation} gets the familiar form of the Stokes equation:
\begin{equation}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = - \rho \mathbf{g}.
\label{eq:stokes}
\end{equation}

For the purposes of this analysis it is useful to define a hydrostatic reference state where the 
velocity $\mathbf{u}$ is zero:
\begin{equation}
\nabla p_0 = \rho_0 \mathbf{g},
\label{eq:hydrostatic_stokes}
\end{equation}
where $p_0$ is the reference hydrostatic pressure and $\rho_0$ is a reference density profile (which may vary with depth).
The total pressure and density can then be decomposed into variations about their 
reference values: $\rho = \rho_0 + \rho^\prime$, $p = p_0 + p^\prime$.
Using this, the hydrostatic reference state (Equation~\eqref{eq:hydrostatic_stokes}) may be subtracted from Equation~\eqref{eq:stokes}.

This gives rise to the following time dependent, coupled system with unknowns $\mathbf{u}(t)$ and $\mathbf{p}(t)$:
\begin{equation}
\begin{aligned}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime &= - \rho^\prime \mathbf{g} \qquad &\text{in } &\Omega(t)\\
\nabla \cdot \mathbf{u} &= 0  \qquad &\text{in } &\Omega(t)
\end{aligned}
\label{eq:final_system}
\end{equation}
defined on the bounded, moving domain $\Omega(t)\subset \mathbb{R}^d$ with boundary $\partial \Omega = \Gamma_0 \cup \Gamma_F$
split into fixed (Dirichlet) and free surface parts $\Gamma_0$ and $\Gamma_F$, respectively.
The boundary conditions are given by
\begin{equation}
\begin{aligned}
\mathbf{u} &= 0 &\qquad \text{on } &\Gamma_0 \\
\mathbf{T}\cdot \mathbf{n} &= 0 &\qquad \text{on } &\Gamma_F.
\end{aligned}
\end{equation}
For the sake of economy, we neglect inhomogeneous stress boundary conditions, though it would be straightforward to include them.
We also defer discussion of free slip boundary conditions.
The domain at time $t$ is defined by advecting a reference configuration $\Omega_0$ by the domain velocity $\mathbf{u}(t)$
\begin{equation}
 \Omega(t) = \Omega_0 + \int_{t_0}^t \mathbf{u}(t) \text{ d}t.
 \label{eq:domain_evolution}
\end{equation}
We defer the discussion of the velocity of the discretized domain to Section~\ref{sec:remeshing}.
Finally, we denote the displacement at time $t$ by $\zeta = \int_{t_0}^t \mathbf{u}_\mathrm{surface}(t)\cdot \mathbf{n} \text{ d}t$, leading to the evolution
equation
\begin{equation}
\frac{\text{d} \zeta}{\text{d}t} = \mathbf{u \cdot \mathbf{n}} \quad \textrm{on  }  \Gamma_F.
\label{eq:surface_evolution}
\end{equation}

@@

\section{Conclusion}
We have analyzed stability of free surface boundary conditions in geodynamic simulations and 
demonstrated the cause of the drunken sailor instability using a normal mode analysis.
This perspective on the problem allowed us to construct an explicit finite difference 
scheme which is first order accurate in time and is unconditionally stable.
The nonstandard finite difference scheme is simple to implement, and 
requires no modifications to the system matrix.
